---
title: "2042 MLM Mini Project (Spring 2020)"
subtitle: "Group 1"
date:  "`r format(Sys.time(), '%B %d %Y')`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```
 
```{r dependencies, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Dependencies -----------------------------------------------------------------
if(!requireNamespace("here"))
  install.packages("here", repos = "https://cloud.r-project.org")
require("here")

if(!requireNamespace("lme4"))
  install.packages("lme4", repos = "https://cloud.r-project.org")
require("lme4")

if(!requireNamespace("lmerTest"))
  install.packages("lmerTest", repos = "https://cloud.r-project.org")
require("lmerTest")
```

## Team members and division of work

Frank Jiang, Lisa Song, Yuyue Hua, Seeun Jang, Tong Jin

```{r set-seed}
set.seed(2042001)
```

### Question 1

You will generate simulated data for a single school with 100 classrooms, each of which has 200 students. 

  a. Outcome for student $i$ in classroom $j$: $Y_{ij}$.
    
  b. There is a single predictor, $X_{ij} \sim U(0,1)$ (uniform on [0,1])
    
  c. There is a classroom random effect, $\eta_j \sim N(0,\sigma^2_{\eta})$, where $\sigma^2_{\eta} = 2$. 
  
  d. Subject level error, $\varepsilon_{ij} \sim N(0, \sigma^2_{\varepsilon})$, where  $\sigma^2_{\varepsilon} = 2$. 
  
  e. `set.seed(2042001)` once at the beginning of your code.
  
  f. Generate the random quantities in this order to ensure the same solution for everyone: X, $\eta_j$, $\varepsilon_{ij}$ 
  
  g. The outcome has the following form (DGP, given the modeling parameters above):
  
$$
Y_{ij} = 
0 + 
1X_{ij} + 
\eta_j + 
\varepsilon_{ij} ;
\mbox{     } 
\eta_j \sim N(0,\sigma^2_{\eta}), 
\varepsilon_{ij} \sim N(0, \sigma^2_{\varepsilon}), 
\text{indep.} 
$$

  h. Generate a single simulated dataset (you will need a "classid" variable to track classrooms); you can optionally assign a "studentid")
    i. **Important:** construct classid such that classrooms appear consecutively within the dataframe. As per: `rep(1:J,each=n_j)`
  
  
```{r Q1}
# Compute variables related to the equation
N_j <- 100 # Total number of classrooms
n_i <- 200 # Number of students in each classroom
N_i <- N_j * 200 # Total number of students

X_ij <- runif( # Single predictor: Uniform on [0, 1]
  N_i, 
  min = 0, 
  max = 1
)
eta_j <- rnorm( # Classroom random effect: Normal on (0, 2)
  N_j,
  mean = 0,
  sd = sqrt(2)
)
epsilon_ij <- rnorm( # Subject level error: Normal on (0, 2)
  N_i,
  mean = 0,
  sd = sqrt(2)
)

# Create equation elements
eta_j <- rep( # Assign classroom random effect to students in each classroom
  eta_j,
  each = n_i
)
Y_ij <- 0 + 1 * X_ij + eta_j + epsilon_ij

classid <- rep(
  1:N_j,
  each = n_i
)
studentid <- seq(
  1:N_i
)

# Create a dataframe to store all elements
classroom_sim <- data.frame(
  outcome = Y_ij,
  predictor = X_ij,
  cls_raneff = eta_j,
  subject_error = epsilon_ij,
  classid = classid,
  studentid = studentid,
  row.names = studentid
)
```

### Question 2

Fit the model corresponding to the DGP on your simulated data.

```{r Q2}
# Fit the model in Q1
fit_q2 <- lmer(
  outcome ~ predictor + (1 | classid),
  data = classroom_sim
)

slope_X <- round(
  summary(fit_q2)$coefficients['predictor', 'Estimate'],
  digits = 4
)
```

a. Report coefficient estimate for slope on X.

    Response: `r slope_X`.

b. Does a 95% confidence band for this coefficient estimate cover the “truth” that you used to generate the data? **comment**

```{r Q2_b}
CI_Q2_b <- confint(
  fit_q2,
  parm = "predictor",
  level = 0.95
)
lower <- round(
  CI_Q2_b[1],
  digits = 4
)
upper <- round(
  CI_Q2_b[2],
  digits = 4
)
```
    Response: Yes. The confidence band (95% of confidence) for this coefficient estimate is between `r lower` and `r upper`. This confirms that a 95% confidence band covers the "truth" of the slope, which is 1.
    
### Question 3

3. Next, we simulate missing data in several ways. This is the first:
a. Make a copy of the data, then modify the copy following these instructions:
```{r}
# Insert code to make a copy of the data
```

b. Generate $Z_{ij} \sim$ Bernoulli$(p)$, with $p = 0.5$
c. Set Y to NA when $Z_{ij} == 1$. This should look a lot like "MCAR" missingness.
```{r}
# Insert code the generate your data
```

d. Refit the model on the new data and report the coefficient estimate for slope on X. Look at the other parameter estimates as well.
```{r}
# Insert code to fit model and compute confidence interval
```

e. Do you see any real change in the $\beta_X$ estimate? **comment**
    i. Does a 95% confidence band for this coefficient estimate cover the “truth” that you used to generate the data?
    
    Response: 

f. What is the total sample size $N$ used in the model fit? **comment**

    Response:
    
### Question 4: 
Missing Data II: Make another copy of the original data, then modify the copy as follows:
a. Generate $Z_{ij} \sim$ Bernoulli$(X_{ij})$, with $X_{ij}$ your predictor generated previously.
b. Set Y to NA when $Z_{ij} == 1$. This should look a lot like "MAR" missingness.
```{r}
# Insert code the generate your data
```
c. Refit the model on the new data and report the coefficient estimate for slope on X. Look at the other parameter estimates as well. **comment**
```{r}
# Insert code to fit model and compute confidence interval
```

    Response: 

d. Do you see any real change in the $\beta_X$ estimate?
    i. Does a 95% confidence band for this coefficient estimate cover the “truth” that you used to generate the data? **comment**
    
    Response:

e. What is the total sample size $N$ used in the model fit? **comment**

    Response:

### Question 5: 
Missing Data III: Make another copy of the original data, then modify the copy as follows:
```{r}
# Insert code to make a copy of the original data
```

a. First, define the expit function: `expit <- function(x) exp(x)/(1+exp(x))`
```{r}
# Insert code to define expit function
```

b. Generate $Z_{ij} \sim$ Bernoulli$(expit(Y_{ij}))$, with $Y_{ij}$ your *outcome* generated previously. 

c. Set Y to NA when $Z_{ij} == 1$. This should look like a violation of "MAR" missingness (missingness depedents on outcome and cannot be *simply* predicted with the predictor set -- Y should be correlated with X, though, so it might not be too bad a violation). 
```{r}
# Insert code the generate your data
```

d. Refit the model on the new data and report the coefficient estimate for slope on X. Look at the other parameter estimates as well. **comment**
```{r}
# Insert code to fit model and compute confidence interval
```

    Response:

e. Do you see any real change in the $\beta_X$ estimate? **comment**
    i. Does a 95% confidence band for this coefficient estimate cover the “truth” that you used to generate the data? **comment**
    
      Response:
    
f. What is the total sample size $N$ used in the model fit? **comment**

    Response:







